---
title: "Analysis of Income, Temperature, and Population Density in Spain 2020"
author: "Nykol Armacanqui and Labanya Paul"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,      # hides code
  warning = FALSE,   # hides warnings
  message = FALSE    # hides messages
)
```

### Setup

This section loads all necessary libraries and sets up the environment for the analysis.

```{r load-libraries}
library(pacman)

pacman::p_load(
  httr,
  jsonlite,exactextractr,stargazer,
  ggplot2,
  keyring,
  ecmwfr,
  dplyr,
  reticulate,
  geodata,
  sf,
  rvest,
  raster,
  rnaturalearth,
  rnaturalearthdata,
  grid,
  readr,
  ggspatial,
  gridExtra,
  ggrepel,
  tidygeocoder,
  osrm ,
  here,
  terra,
  tidyr,
  tmap,
  ineAtlas,
  tmap
)
here::i_am("assignment_3.Rmd")  
```

### API and Data Configuration

Here, we configure the download folder and set up the API key for the Copernicus Climate Data Store (CDS) to download temperature data.

```{r api-setup}
# download folder setup
download_folder <- here("data", "raw", "raster")

# Make sure the folder exists
if (!dir.exists(download_folder)) dir.create(download_folder, recursive = TRUE)

# key setup
key_set_with_value(
  service  = "cds",
  username = "nykol_yove.armacanqui_martinez_de_tudor@mailbox.tu-dresden.de",
  password = "3aa6e68c-6a56-4adb-a9c4-2e6220ac9806"
)

# now pull it back out:
cds_api_key <- key_get(
  service  = "cds",
  username = "nykol_yove.armacanqui_martinez_de_tudor@mailbox.tu-dresden.de"
)


# Export to environment variables for cdsapi, reticulate, or ecmwfr
Sys.setenv(
  CDSAPI_URL = "https://cds.climate.copernicus.eu/api",
  CDSAPI_KEY = cds_api_key
)


wf_set_key(
  key  = cds_api_key
)


# if you ever want to remember your API key 
wf_get_key()
```

### Data Download: ERA5 Temperature


```{r download-era5, eval=FALSE}
# This chunk is set to eval=FALSE to prevent re-downloading the data on every run.
# Set to TRUE if you need to download the data for the first time.
request <- list(
  dataset_short_name = "reanalysis-era5-single-levels",
  product_type       = "reanalysis",
  variable           = "2m_temperature",
  year               = "2020",
  month              = "07",
  day                = "01",
  time               = "15:00",
  data_format        = "netcdf",
  download_format    = "unarchived",
  target             = "era5_temp_20200107.nc"
)

era5 <- wf_request(
  request  = request,
  transfer = TRUE,
  path     = download_folder
)
```

### Vector Data: Provincial Income

We will now load and process the income data. This involves reading a CSV file, aggregating municipal income to the provincial level, and joining it with spatial data for Spain's provinces.

```{r process-income-data}

#income_municipalities csv file
income_municipalities <- read_csv(here("data", "raw", "vector", "income_mun.csv"))

spain_prov <- geodata::gadm(country = "ESP", level = 2, path = tempdir())

#checking 
plot(spain_prov["NAME_2"], main = "Spain Provinces")

#filter to year of interest 

income_2020 <- income_municipalities %>%
  filter(year == 2020)

# Aggregate mean income by province
prov_income <- income_2020 %>%
  group_by(prov_name) %>%
  summarise(mean_income = mean(net_income_pc, na.rm = TRUE))

# checking 

head(prov_income)

unique(spain_prov$NAME_2)
unique(prov_income$prov_name)

# Convert SpatVector to sf object
spain_prov_sf <- sf::st_as_sf(spain_prov)
```

### Joining Spatial and Income Data

Now we join the aggregated income data with the provincial boundaries.

```{r join-income-spatial}

# Load GADM data for Spain
gadm_spain <- geodata::gadm(country = "ESP", level = 2, path=tempdir())
gadm_spain <- st_as_sf(gadm_spain)  # Convert to sf object
gadm_spain <- st_transform(gadm_spain, crs("EPSG:4326"))


# Joining the income data with the spatial data GADM

spain_prov_income <- gadm_spain %>%
  left_join(prov_income, by = c("NAME_2" = "prov_name"))
```

### Map 1: Average Provincial Income

This map visualizes the average net income per capita for each province in Spain for the year 2020.

```{r map-income}

tm_shape(spain_prov_income) +
  tm_fill(
    fill = "mean_income",
    fill.scale = tm_scale(values = "brewer.greens"),
    fill.legend = tm_legend(title = "Mean Income (€)")
  ) +
  tm_borders(col = "black", lwd = 0.5) +
  tm_title("Average Provincial Income in Spain (2020)")
```

### Raster Data: Processing Temperature

Next, we load the downloaded temperature data, reproject it to match our vector data's CRS, and join it to the boundaries of Spain.

```{r process-temperature-raster}

#Getting temperature data from copernicus

# Load the temperature raster
temp_tiff <- rast(here("data", "raw", "raster", "era5_temp_20200107.nc"))

# Reproject temp_tiff to EPSG:4326
temp_tiff_4326 <- project(temp_tiff, crs("EPSG:4326"))

# Clip temperature raster to Spain's extent
temp_spain <- crop(temp_tiff_4326, spain_prov_sf)
temp_spain <- mask(temp_spain, spain_prov_sf)

# Plot to check
plot(temp_spain, main = "Temperature in Spain (July 1, 2020 at 15:00)")
```

### Extracting Temperature and Filtering Data

We extract the mean temperature for each province and then filter the dataset to include only mainland Spain, removing the Canary Islands.

```{r extract-temp-and-filter}
# mean temperature per province
spain_prov_income$mean_temp <- exact_extract(temp_spain, spain_prov_income, "mean")

# Check result
head(spain_prov_income[, c("NAME_2", "mean_income", "mean_temp")])

# eliminating "Islas Canarias"from the map border to just be mainland (it makes it clearer)

spain_prov_income <- spain_prov_income %>%
  filter(NAME_1 != "Islas Canarias")
```

### Map 2: Average Provincial Temperature

This map shows the average temperature across mainland Spanish provinces. 

```{r map-temperature}
# Plotting the temperature data 
tm_shape(spain_prov_income) +
  tm_fill(
    fill = "mean_temp",
    fill.scale = tm_scale(values = "matplotlib.reds"),
    fill.legend = tm_legend(title = "Mean Temperature (°C)")
  ) +
  tm_borders() +
  tm_title("Average Provincial Temperature in Spain in 2020 (Mainland Spain)")
```

### Analysis 1: Simple Linear Regression (Income vs. Temperature)

We first look at the relationship between income and temperature with a scatterplot and then run a simple linear regression.

```{r scatterplot-income-temp}
##Scatterplot the relationship between income and temperature

ggplot(spain_prov_income, aes(x = mean_temp, y = mean_income)) +
geom_point(color = "darkblue", size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Income vs Temperature in Spain",
       x = "Mean Temperature (°C)",
       y = "Mean Income (€)")
```

```{r simple-regression}
#Dropping NA values
spain_prov_income <- spain_prov_income %>%
  filter(!is.na(mean_income) & !is.na(mean_temp))

# regression
Regression  <- lm(mean_income ~ mean_temp, data = spain_prov_income)
# Display regression in a table
stargazer(Regression, type = "text", title = "Regression: Income vs Temperature")
# Display regression summary
summary(Regression)
```

### Raster Data: Processing Population Density

To improve our model, we introduce population density as a control variable. We load, process, and log transform the population density data.

```{r process-population-raster}
####################################################### Raster
#### We also have the Population density 

# Loading the population density raster
popd_data <- rast(here("data", "raw", "raster", "esp_pd_2020_1km.tif"))
# Ensure popd_data is in EPSG:4326
if (crs(popd_data) != "EPSG:4326") {
  popd_data <- project(popd_data, crs("EPSG:4326"))
}

## Cropping for España
spain <- ne_countries(scale = "medium", returnclass = "sf") %>%
  filter(admin == "Spain")
spain <- st_transform(spain, crs("EPSG:4326"))

# Crop and mask population density raster to Spain
popd_spain <- crop(popd_data, vect(spain))
popd_spain <- mask(popd_spain, vect(spain))


# log the population density raster (doing this to make it look more clear)

popd_spain <- log(popd_spain + 1)  # Adding 1 to avoid log(0)


# plot with x and y limits (mainland España)

plot(popd_spain, main = "Log Population Density in Spain (2020)",
     xlim = c(-10, 5), ylim = c(35, 45))
```

### Extracting Population Density and Finalizing Data

We extract the mean population density for each province and clean the dataset of any remaining NA values.

```{r extract-popd}
# mean population density to province dataset
spain_prov_income$mean_popd <- exact_extract(
  popd_spain, spain_prov_income, "mean"
)

# Check
head(spain_prov_income[, c("NAME_2", "mean_income", "mean_temp", "mean_popd")])

# Now each province has mean income, mean temperature, and mean population density.

# Dropping NA values
spain_prov_income <- spain_prov_income %>%
  filter(!is.na(mean_income) & !is.na(mean_temp) & !is.na(mean_popd))
```

### Analysis 2: Simple Linear Regression (Income vs. Population Density)

A scatterplot helps show the relationship between provincial income and population density.

```{r scatterplot-income-popd}
# plot (line) population density and mean income

ggplot(spain_prov_income, aes(x = mean_popd, y = mean_income)) +
  geom_point(color = "darkgreen", size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Income vs Population Density in Spain",
       x = "Mean Population Density (log scale)",
       y = "Mean Income (€)") +
  scale_x_log10()  
```

### Preparing Data for Final Maps

We create new columns to label income and identify the top 3 richest and bottom 3 poorest provinces.

```{r prep-map-labels}
# top 3 provinces by income
spain_prov_income <- spain_prov_income %>%
  mutate(
    income_label = paste0("€", signif(mean_income / 1000, 3), "k"),
    top3 = rank(-mean_income) <= 3
  )

# top 3 richest 
spain_prov_income %>% filter(top3)

# checking the lowest mean income provinces 
spain_prov_income <- spain_prov_income %>%
  mutate(bottom3 = mean_income %in% sort(mean_income, na.last = NA)[1:3])

# Check bottom 3 provinces
spain_prov_income %>% filter(bottom3)
```

### Map 3: Population Density and Income

This map overlays mean income labels on population density. The top 3 richest provinces are highlighted in bold red.

```{r map-popd-income}
tm_shape(spain_prov_income) +
  tm_fill(
    fill = "mean_popd",
    fill.scale = tm_scale(values = "brewer.yl_gn_bu"),
    fill.legend = tm_legend(title = "Mean Population Density (log)")
  ) +
  tm_borders(col = "black", lwd = 0.5) +
  tm_shape(spain_prov_income %>% filter(!top3)) +
  tm_text("income_label", size = 0.5, col = "black", shadow = TRUE) +
  tm_shape(spain_prov_income %>% filter(top3)) +
  tm_text("income_label", size = 0.5, col = "red", fontface = "bold", shadow = TRUE) +
  tm_title("Mean Population Density and Income in Mainland Spain (2020)", size = 0.7) +
  tm_layout(
    legend.outside = TRUE,
    frame = FALSE,
    bg.color = "white",
    legend.title.size = 0.5,   
    legend.text.size = 0.5     
  )
```

### Map 4: Temperature and Income

This map overlays mean income labels on mean temperature. The top 3 richest provinces are highlighted in blue, and the bottom 3 are in green.

```{r map-temp-income}

tm_shape(spain_prov_income) +
  tm_fill(
    fill = "mean_temp",
    fill.scale = tm_scale(values = "brewer.yl_or_rd"),
    fill.legend = tm_legend(title = "Mean Temperature (°C)")
  ) +
  tm_borders(col = "black", lwd = 0.5) +
  tm_shape(spain_prov_income %>% filter(!top3)) +
  tm_text("income_label", size = 0.5, col = "black", shadow = TRUE) +
  tm_shape(spain_prov_income %>% filter(top3)) +
  tm_text("income_label", size = 0.5, col = "blue", fontface = "bold", shadow = TRUE) +
  tm_shape(spain_prov_income %>% filter(bottom3)) +
  tm_text("income_label", size = 0.5, col = "darkgreen", fontface = "bold", shadow = TRUE) +
  tm_title("Mean Temperature and Income in Mainland Spain (2020)", size = 0.7) +
  tm_layout(
    legend.outside = TRUE,
    frame = FALSE,
    bg.color = "white",
    legend.title.size = 0.5,   
    legend.text.size = 0.5     
  )		

```



### Analysis 2: Multiple Linear Regression

Finally, we run a multiple linear regression model to assess the relationship between income and temperature while controlling for population density. We then compare this model to our earlier simple regression.

```{r multiple-regression}
####Running a linear regression model, with population density as a control variable
Regression_pop <- lm(mean_income ~ mean_temp + mean_popd, data = spain_prov_income)


# Display regression in a table
stargazer(Regression_pop, type = "text", title = "Multiple Regression: Income vs Temperature and Population Density")
# Display the two regression summaries in a table
stargazer(Regression, Regression_pop, type = "text", 
          title = "Comparing Simple OLS and OLS with POP Control",
          column.labels = c("Simple OLS", "Control for Population Density"),
          dep.var.labels = "Mean Income (€)",
          covariate.labels = c("Mean Temperature (°C)", "Mean Population Density (people/km²)"),
          model.numbers = FALSE)
```
```{r}

```
